Data Specialization for a Staged Lambda Calculus

### 2015-02-18 paper/semantics.tex
one figure per file
almost done messing with figures...
finished a pass through 3.3

\input{figures/splitting-overview}

@@

@@

$\next{e}$ is trickier, because we should not evaluate $e$ (at \bbtwo) yet.
There are two difficulties---$e$ may contain other code at \bbonem\ (which we
handle with $\redtwosym$, discussed below); and $\next{e}$ may be part of a
first-stage redex, which we \emph{do} need to reduce. For example, in
\begin{lstlisting} 
1`(fn x : $int => e') (next{`2`e`1`})`
\end{lstlisting}
we must reduce this first-stage function application, but cannot reduce $e$ yet.
We could reduce this to $[\next{e}/x]e'$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to hoist $e$ outside, binding it
to a temporary variable $\hat{y}$, and substituting that variable instead:
\begin{lstlisting} 
2`let yhat = e in `1`[next{`2`yhat`1`}/x]e'`
\end{lstlisting}
then proceed by evaluating $[\next{\hat{y}}/x]e'$.

@@

\subsection {Top-Level Evaluation}

\crem{TODO below here}

We can use the evaluation relations in the dynamic semantics to
evaluate a term to evaluate a \lang\ term as follows. We first apply
1-evaluation to obtain a partial value and a residual table. We then
construct a 2-term from the partial value and the residual table by
using reification, and then reduce this term to a residual via
residualization.  Finally, we evaluate the residual, which is a term in a
our monostaged target language \langmono, to a value by using  a
standard evaluation relation for \langmono.

We can evaluate a \lang\ term as
follows: we first apply 1-evaluation to obtain a partial value and a
residual table, we then construct a 2-term from the partial value and
the residual table by using reification, and then reduce this term to
a residual via residualization.

These dynamics allow us to define a partial evaluator for \lang\ by identifying
static with stage \bbone\ and dynamic with stage \bbtwo.  Specifically, we
encode $f$ as a \lang\ expression with a function type of the form
$A\to\fut(B\to C)$.%

Here $A$ is the static input, $B$ is the dynamic input, and $C$ is the output.

Once a stage \bbone\ argument $a:A$ is provided, we can evaluate the partially-applied
function:
$\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
C)$, which by canonical forms must have the form $v = \next~\hat y$. 
Next, we reify this environment into a sequence of \verb|let|-bindings
enclosing $\hat y$, via $\reify\xi{\hat y}{f_a}$. 
Because reification preserves types, the resulting residual $f_a$ has type $B\to C$ in \langTwo, so we can apply it to some $b:B$
and compute the final result of the function, $f_a~b \mathop{\tworedsym} c$.

That this sequence of evaluations is in fact staged follows from our
characterizations of partial values and residuals, that $\redonesym$
outputs a partial value, and that $\reifysym$ outputs an expression in \langTwo.

For example, to specialize quickselect to a particular list, 
\begin{lstlisting}
qsStaged [5,2,7,4,1]
\end{lstlisting}
\TODO finish example

%We can rewrite \texttt{fexp} in this form, or simply apply
%the following higher-order function which makes the adjustment:
%\begin{lstlisting} 
%let adjust (f : $int * int -> $int) =
%  fn (p : int) => 
%    next{
%      fn (b : int) => 
%        prev{f (next {b}, p)}
%    }
%\end{lstlisting}

>>>>

\input{figures/values}

@@

\input{figures/dynamics-overview}

@@

$\next{e}$ is trickier, because we should not evaluate $e$ (at \bbtwo) yet.
There are two difficulties---$e$ may contain other code at \bbonem\ (which we
handle with $\redtwosym$, discussed below); and $\next{e}$ may be part of a
first-stage redex, which we \emph{do} need to reduce. For example, in
\begin{lstlisting} 
1`(fn x : $int => e') (next{`2`e`1`})`
\end{lstlisting}
we must reduce this first-stage function application, but cannot reduce $e$ yet.
We could reduce this to $[\next{e}/x]e'$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to hoist $e$ outside, binding it
to a temporary variable $\hat{y}$, and substituting that variable instead:
\begin{lstlisting} 
2`let val yhat = e in `1`[next{`2`yhat`1`}/x]e'`
\end{lstlisting}
then proceed by evaluating $[\next{\hat{y}}/x]e'$.

@@

\subsection{Evaluating Staged Programs}

We consider a staged program to be a closed term $e$ at world \bbtwo. The first
stage of evaluation is $e\mathbin{\redtwosym} q$, which reduces all code at
\bbonem\ and \bbonep\ inside $e$. The second stage of evaluation is $q
\mathbin{\redsym} u$, or monostaged evaluation of the residual $q$, which yields
a value $u$.

A multistage function, such as \texttt{qsStaged} from \ref{sec:staging}, has
type $A\to \fut (B \to C)$ at \bbonem.%
\footnote{We can rewrite \texttt{qsStaged} in this curried form, or apply a
higher-order currying function.}
To apply this function to arguments $\colmix a A$ and $\coltwo b B$, simply
evaluate the program:
\begin{lstlisting}
2`prev{`1`f a`2`} b`
\end{lstlisting}

We can also directly observe the reuse of first-stage computations across
multiple second-stage computations. To run many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
\begin{lstlisting}
2`let val qs = prev{`1`qsStaged `3`grnd{`1`[7,4,2,5,9,...,3]`3`}`2`}`
2`in (qs k1,...,qs km)`
\end{lstlisting}
To see that the partial application of \texttt{qsStaged} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let val yhat = ...`
2`    val qs = yhat`
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $\hat{y}$ is bound to a function containing no first-stage redexes.
